Optimal Finite Element Integration

### 2015-08-07 TOMS_Paper/assembly.tex
Minor fixes

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

@@

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available (\cite{ufl-code}).

@@

Also note some other minor things.
\begin{itemize}
\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of the hyperelastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. We note how the fact that\texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present. With $q=4$, pre-evaluation is triggered in \texttt{cfO2} when $nf0$. When $nf=1$, pre-evaluation is found profitable in terms of operation count, but it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$. We have tried running the same experiments with $T_H = \infty$ and noticed a significant drop in runtime for all values of $p$ (\texttt{cfO2} was generally slower than \texttt{cfO1}). This confirms the importance of memory constraints.

>>>>

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

@@

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.

@@

Also note some other minor things.
\begin{itemize}
\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. We note how the fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present. With $q=4$, pre-evaluation is triggered in \texttt{cfO2} when $nf0$. When $nf=1$, pre-evaluation is found profitable in terms of operation count, although it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$. We have tried running the same experiments with $T_H = \infty$ and noticed a significant drop in runtime for all values of $p$ (\texttt{cfO2} was generally slower than \texttt{cfO1}). This confirms the importance of memory constraints.
